---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,echo=FALSE}
options(scipen = 6)
```

## Loading and preprocessing the data


The variables included in this dataset are:  

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken


The data is loaded from activity.csv and stored in activity. 

```{r loaddata,echo=TRUE}
activity <- read.csv("activity.csv") 
```

## What is mean total number of steps taken per day?


###1. The total number of steps taken per day###

The total number of steps taken each day is calculated using the code below. The steps taken on each day in the data, where a valid reading is recorded (not NA) are summed and recorded in days.Steps. 

```{r daysteps,echo=TRUE}
library(plyr)
# Sum steps for each day where the value is recorded for the interval
days.Steps <- ddply(subset(activity, !(is.na(steps)) ), "date", colwise(sum))
```

The following cade prints the table containing the calculated steps taken on each day (contents of days.Steps). The output shows the date and the total steps taken on that date.   

```{r displayframe,echo=TRUE,results = 'asis',warning=FALSE}
library(knitr)
kable(days.Steps[,1:2], align="c")
```
   
###2. The data in days.Steps is used to create a histogram of the total number of steps taken each day ###   

The code below breaks the range of steps in to 5 and plots a histogram of the frequency of each of the 5 range steps in terms of days.

```{r plotstepshist, echo=TRUE}
hist(days.Steps$steps, breaks=5, main="Steps Taken Each Day", xlab = "Steps taken", ylab="Number of days (frequency)")
```


###3. The mean and median of the total number of step taken per day (in days.Steps) are then calcuated###

The code below calculates the mean and median steps per day.   
**Note**: where no measurements were available in the activity data for an entire day, the day is excluded from the calculation of the mean.   

```{r meanstepsperday,echo=TRUE}
mean.Steps<-apply(days.Steps[2], 2, mean)
median.Steps<-apply(days.Steps[2], 2, median)
```


Using the calculated mean and median values above: 

- The **mean** total number of steps taken per day (mean.Steps) is calculated to be **`r mean.Steps`**.   
- The **median** total number of steps taken per day (median.Steps) is calculated to be **`r median.Steps`**.   

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
