---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,echo=FALSE}
options(scipen = 6)
```

## Loading and preprocessing the data


The variables included in this dataset are:  

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken


The data is loaded from activity.csv and stored in activity.

```{r loaddata,echo=TRUE}
activity <- read.csv("activity.csv") 
```

## What is mean total number of steps taken per day?

The code below is used to add up the steps taken on each day in the data, where a reading is recorded (not NA).     
**Note**: where no measurements are available, the day is excluded from the calculation of the mean.   

```{r meanstepsperday,echo=TRUE}
library(plyr)
# Sum steps for each day where the value is recorded for the interval
days.Steps <- ddply(subset(activity, !(is.na(steps)) ), "date", colwise(sum))
# Calculate the mean steps over the days in the data
mean.Steps<-apply(days.Steps[2], 2, mean)
```

Using the above code the **mean total number of steps taken per day** (mean.Steps) is calculated to be **`r mean.Steps`**.   

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
